Facility and Time Step Discretization Effects in Fuel Cycle Simulation

### 2015-09-29 paper.tex
edits up to scenario description

Due to the diversity of fuel cycle simulators' modeling assumptions direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps one of the most complete published comparisons performed.  Despite
this, however, various results from the simulators were often significantly
different due to inconsistency in modeling decisions such as reprocessing
strategies, refueling behavior, reactor end-of-life handling, etc.  This work
attempts to identify and quantify the effects of a few different modeling
choices that may sometimes be taken for granted in the fuel cycle modeling
domain.  Four scenarios were compared using combinations of fleet-based or
individual reactors with monthly or quarterly (3-month) time steps.  When
under supply-constraint pressure, the four cases exhibit noticably different
behavior.  Otherwise, they generally behaved very similarly.  Fleet-based
modeling is more efficient in supply-constrained environments at the expense
of losing insight on issues such as reactor refueling cycle staggering.
Finer-grained time steps also enable more efficient material use in
supply-constrained environments. Large simulations with fleet-based reactors
run much more quickly than their individual reactor counterparts.

@@

Due to the diversity of fuel cycle simulators' modeling assumptions direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study
\cite{oecd2012benchmark} that is perhaps one of the most complete published
comparisons performed.  Despite this, however, various results from the
different simulators were often significantly different due to different
modeling decisions such as reprocessing strategies, refueling behavior,
reactor end-of-life handling, etc.  This work attempts to identify and
quantify the effects of a few different modeling choices that may sometimes be
taken for granted in the fuel cycle modeling domain: facility modeling
discretization (i.e. individual facilities vs. aggregate fleets) and time step
duration.  A better understanding of how these affect fuel cycle dynamics can
allow for making more deliberate decisions with respect to tradeoffs such as
performance vs. realism.

@@

Traditionally, many fuel cycle simulators have used a system dynamics approach
\cite{forrester_industrial_1961} to modeling. These simulators model several
"stocks" that have levels that change over time due to "flows" between them.
The flows are determined by an equation that is a function of the state of the
system (i.e. the system's stocks).  A constructed system is then solved
in discrete or continuous time depending on the limitations of the software
and constraints imposed by the model.  In system dynamics simulators, the
number of stocks and potential flows between them is generally a static
property of the modeled system, making it more difficult to swap different
reactor models in and out of the same scenario without wiring them into the
simulator manually.

System dynamics based fuel cycle simulators generally use a collection of
stocks to represent the state of groups of like facilities commonly referred
to as "fleets".  As the simulation steps through time, the levels of stocks
(e.g. reactor fresh fuel inventory, repository waste inventories, etc.) are
adjusted according to the calculated flow values.  Simulators that model
facility fleets built on a system dynamics foundation include VISION
\cite{jacobson_verifiable_2010}, DANESS \cite{van_den_durpel_daness_2009},
DYMOND (predecessor to DANESS), and ORION \cite{gregg_orion_2012} among
others.  In fleet-based modeling:

@@

This work investigates in detail differences between fleet vs. individual
reactor modeling.  The difference between these two modeling styles in terms
of their affects on output of fuel cycle simulators in general is currently
unknown. This analysis is designed to provide both quantitative and
qualitative insight about how fleet-based continuous flow models differ from
individual facility discrete flow models.

Many fuel cycle simulators step through time in discrete steps.  The duration
of a single time step can have significant impact on the dynamics of a
simulation.  The time step duration can affect the magnitude of individual
material transfers between facilities and correspondingly affects the noise
level in facility material inventories.  It can also act as a minimum bound
for outage durations. The effects of the time step duration will also be
investigated together with fleet vs facility modeling.

@@

As a Cyclus simulation walks through each time step, facilities engage in a
market-like dynamic resource exchange (DRE).  They broadcast requests for
material and suppliers respond with bids. The DRE then creates a network flow
problem that is optimized to resolve which transactions actually occur.  The
plugin architecture combined with the DRE creates a powerful and flexible
environment for the natural comparison of different modeling paradigms.
Cyclus' agent-based architecture in concert with its dynamic resource exchange
has the ability to model fleet-based facilities.  However, Cyclus was designed
to enable modeling facilities individually with discrete flows enabling the
investigation of interesting real-world effects such as competition, reactor
cycle staggering, and individual reactor outages.

Cyclus was designed to operate natively in unix/linux type environments with
flexible ascriptable usage and input file format.  Cyclus holds its diverse
generated data into a powerful open-source database format.  In addition to
information about material flows and facility deployments, Cyclus' single-file
database contains version information about the archetypes used in a
simulation, the simulation input file, version of Cyclus used to run the
simulation, and the installed versions of Cyclus dependencies. This and other
features of Cyclus lend themselves well to automation, research provenance and
reproducibility, and large-scale computing.

@@

Because the standard package included with Cyclus does not include a
fleet-based reactor model, one was necessarily developed. A simulation
scenario was created that is roughly based on prior and ongoing work in the
Department of Energy (DOE).  A few Phenomena related to the different modeling
choices in each of the four cases are identified and used as a basis for
comparison.  Each of these three things is described in more detail in the
following sections.

@@

A fleet reactor facility was created for Cyclus.  The flexibility of Cyclus
makes it relatively straight forward to create plug-in models for Cyclus that
live anywhere along the spectrum from individual, discrete facilities to
continuous flow fleet facilities.  In order to contrast well with individual,
single-facility granularity modeling, the fleet model created here falls more
toward the aggregate, continuous flow side of the spectrum.

The fleet reactor, although a single entity modeling many reactors, is
actually designed to look like many individual reactors to the cyclus
simulator kernel. This "trick" allows other plugins for managing facility
deployments to transparently work with the fleet reactor without having to
know anything about the fleet concept.  The fleet reactor has several
important characteristics described below:

@@

        $D$ is the discharge rate (kg per time step), $B$ is the batch size
        (kg) for a single reactor's batch, $L$ is the cycle length (including
        refueling time) in time steps, $N_{r}$ is the number of
        reactors in the fleet, $C_{inv}$ is the total amount of fuel in all
        reactor cores (kg), and $C_{cap}$ is the total fuel capacity for all
        reactor cores (kg).  The last fraction term accounts for any amount of
        underfueling that may be occuring resulting in lower fuel burning
        rates due to unfueled reactors.

    \item Refueling occurs continuously with as much fresh fuel as is
        available up to the total amount to fill all reactor cores in the
        fleet.

    \item Generated power is computed similarly to core discharge rates by
        multiplying the total fleet power capacity by the same last term of
        Equation \ref{eqn:fleet-discharge} to account for unfueled capacity in
        the fleet.

@@

These characteristics result in a fleet reactor having perfect fuel sharing
and perfect cycle staggering among the individual reactors it models.  First
discharge of spent fuel from fleets begins immediately rather than after a
complete refueling cycle. Fuel shortages also only cause proportional loss of
capacity rather than whole-reactor quantized shutdown.  The full
implementation of the fleet-based reactor is publicy available for download
\cite{Carlsen2015}.

>>>>

Due to the diversity of fuel cycle simulators' modeling assumptions direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps one of the most complete published comparisons performed.  Despite
this, however, various results from the simulators were often significantly
different due to inconsistency in modeling decisions involving reprocessing
strategies, refueling behavior, reactor end-of-life handling, etc.  This work
attempts to identify and quantify the effects of a few different modeling
choices that may sometimes be taken for granted in the fuel cycle modeling
domain.  Four scenarios were compared using combinations of either fleet-based
or individual reactors with either monthly or quarterly (3-month) time steps.
When under supply-constraint pressure, the four cases exhibit noticably
different behavior.  Otherwise, they generally behaved very similarly.
Fleet-based modeling is more efficient in supply-constrained environments at
the expense of losing insight on issues such as realistic sub-optimal fuel
distribution and challenges in reactor refueling cycle staggering.
Finer-grained time steps also enable more efficient material use in
supply-constrained environments.  Large simulations with fleet-based reactors
run much more quickly than their individual reactor counterparts.

@@

Due to the diversity of fuel cycle simulators' modeling assumptions direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study
\cite{oecd2012benchmark} that is perhaps one of the most complete published
comparisons performed.  Despite this, however, various results from the
different simulators were often significantly different due to inconsistent
modeling decisions involving reprocessing strategies, refueling behavior,
reactor end-of-life handling, etc.  This work attempts to identify and
quantify the effects of a few different modeling choices that may sometimes be
taken for granted in the fuel cycle modeling domain: facility modeling
discretization (i.e. individual facilities vs. aggregate fleets) and time step
duration.  A better understanding of how these affect fuel cycle dynamics can
allow for making more deliberate decisions with respect to tradeoffs such as
performance vs. realism.

@@

Traditionally, many fuel cycle simulators have used a system dynamics approach
\cite{forrester_industrial_1961} to modeling. These simulators model several
"stocks" representing isotopes or isotope groups that have levels that change
over time due to "flows" between them.  The flows are determined by equations
that are a function of the state of the system (i.e. the system's stocks).  A
constructed system is then solved in discrete or continuous time depending on
the limitations of the software and constraints imposed by the model.  In
system dynamics simulators, the number of stocks and potential flows between
them is generally a static property of the modeled system, making it somewhat
difficult to swap different reactor models in and out of the same scenario
without wiring them into the simulator manually.

System dynamics based fuel cycle simulators generally use a collection of
stocks to represent the state of groups of like-facilities commonly referred
to as "fleets".  As the simulation steps through time, the levels of stocks
(e.g. reactor fresh fuel inventory, repository waste inventories, etc.) are
adjusted according to the calculated flow values.  Simulators that model
facility fleets built on a system dynamics foundation include VISION
\cite{jacobson_verifiable_2010}, DANESS \cite{van_den_durpel_daness_2009},
DYMOND (predecessor to DANESS), and ORION \cite{gregg_orion_2012} among
others.  In fleet-based modeling:

@@

This work investigates in detail some differences between fleet vs. individual
reactor modeling.  The difference between these two modeling styles in terms
of their affects on output of fuel cycle simulators in general is currently
unknown. This analysis is designed to provide both quantitative and
qualitative insight about how fleet-based continuous flow models differ from
individual facility discrete flow models.

Many fuel cycle simulators step through time in discrete steps.  The duration
of a single time step can have significant impact on the dynamics of a
simulation.  The time step duration can affect the magnitude of individual
material transfers between facilities and correspondingly affects the noise
level in facility material inventories and flows.  It can also act as a
minimum bound for outage durations. The effects of the time step duration will
also be investigated together with fleet vs facility modeling.

@@

As a Cyclus simulation walks through each time step, facilities engage in a
market-like dynamic resource exchange (DRE).  They broadcast requests for
material and suppliers respond with bids. Requesters then express preferences
between all their bids, and the DRE then creates a network flow problem that
is optimized to resolve which transactions actually occur.  The plugin
architecture combined with the DRE creates a powerful and flexible environment
for the natural comparison of different modeling paradigms.  Cyclus'
agent-based architecture in concert with its DRE has the ability to model
fleet-based facilities.  Additionally, however, Cyclus can model facilities
individually with discrete flows enabling the investigation of interesting
real-world effects such as competition, reactor cycle staggering, and
individual reactor outages.

Cyclus was designed to operate natively in unix/linux type environments with
flexible and scriptable usage and input file format.  Cyclus holds its diverse
generated data into a powerful open-source database format.  In addition to
information about material flows and facility deployments, Cyclus' single-file
database contains version information about the archetypes used in a
simulation, the simulation input file, version of Cyclus used to run the
simulation, and the installed versions of Cyclus dependencies. This and other
features of Cyclus lend themselves well to automation, research provenance and
reproducibility, and large-scale computing.

@@

Because the standard package of archetypes included with Cyclus does not
include a fleet-based reactor model, one was necessarily developed. A
simulation scenario was created that is roughly based on prior and ongoing
work in the Department of Energy (DOE) fuel cycle options campaign.  A few
phenomena related to the different modeling choices in each of the four cases
are identified and used as a basis for comparison.  Each of these things is
described in more detail in the following sections.

@@

A fleet reactor facility was created for Cyclus.  The flexibility of Cyclus
makes it relatively straight forward to create plug-in archetypes that
live anywhere along the spectrum from individual, discrete facilities to
continuous flow fleet facilities.  In order to contrast well with individual,
single-facility granularity modeling, the fleet model created here falls more
toward the aggregate, continuous flow side of the spectrum.

The fleet reactor, although a single entity modeling many reactors, is
actually designed to look like many individual reactors to the cyclus
simulator kernel. This "trick" allows other plugins for managing facility
deployments to transparently work with the fleet reactor without having to
know anything about the fleet paradigm.  The fleet reactor has several
important characteristics described below:

@@

        $D$ is the discharge rate (kg per time step), $B$ is the batch size
        (kg) for a single reactor's batch, $L$ is the cycle length (including
        refueling time) in time steps, $N_{r}$ is the number of reactors in
        the fleet, $C_{inv}$ is the total amount of fuel in all reactor cores
        (kg), and $C_{cap}$ is the total fuel capacity for all reactor cores
        (kg).  The last fractional term accounts for any amount of
        underfueling that may be occuring resulting in lower fuel burning
        rates due to unfueled reactors.

    \item Refueling occurs continuously with as much fresh fuel as is
        available up to the total amount to fill all reactor cores in the
        fleet.  No extra fuel is kept on-hand - it is all ordered
        just-in-time.

    \item Generated power is computed similarly to core discharge rates by
        multiplying the total fleet power capacity by the same last fractional
        term of Equation \ref{eqn:fleet-discharge} to account for unfueled
        capacity in the fleet.

@@

One characteristic of this fleet reactor implementation is that it has perfect
fuel sharing and perfect cycle staggering among the individual reactors it
models.  The first discharge of spent fuel from fleets begins immediately on
the first time step of operation rather than after a complete refueling cycle.
Fuel shortages also only cause proportional loss of capacity rather than
whole-reactor quantized shutdown.  The full implementation of the fleet-based
reactor is publicy available for download and use \cite{Carlsen2015}.
